---
title: "Homework Week 4: Law of Large Numbers & Central Limit Theorem"
author: "Andrew Joyce"
date: "09-24-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
---

```{r}
#| echo: false
#| results: hide
#| include: false
library(conflicted)
conflicts_prefer(dplyr::filter)
library(tidyverse)
theme_set(theme_light(base_family = "Avenir Next Condensed"))
```

## 4.1

```{r}
x <- rnorm(100000, mean = 0, sd = 1)
# Use the mean() function to verify that x exhibits those four behaviors.

    #1: The probability that x is within one standard deviation away from the mean is roughly 68%.

mean(x >= -1 & x <= 1) 

    #2: The probability that x is within two standard deviations away from the mean is roughly 95%. 

mean(x >= -2 & x <= 2)

    #3. The probability that x is above the mean (>=) is 50%.

mean(x >= 0)

    #4. The probability that x is below the mean (<=) is also 50%.

mean(x <= 0)

```

## 4.2

```{r}
quantile(x)
```

The function `quantile(x)` will show values of x that fall in a certain quantile range (defaults are 0%, 25%, 50%, 75%, and 100%- the extremes \[0%, 100%\] and quartiles \[25%, 50%, 75%\]). 0% of the values are less than -4.027. 25% of the values are less thanqu -0.674. 50% of the values are less than 0.000 (this is the median, which is equivalent to the mean in a normal distribution). 75% of the values are less than 0.681. And finally, 100% of the values are 4.543 or less.

## 4.3

```{r}
# Now modify the probs argument in the quantile() to find the 0.5% and 99.5% percentiles of x.

quantile(x, probs = c(.005, .995))

```

## 4.4

```{r}
# Use the mean() function to verify that the probability of x being between -2.576 and 2.576 is roughly 99%.

mean(x>-2.576 & x<2.576)
```

## 4.5

```{r}
# In R, create 1000 simulations of x (x being the sum of 20 independent, uniform variables) and plot their histogram.

sumx_sim <- tibble(
  sim_num = 1:100) |> 
  rowwise()  |> 
  mutate(sums = sum(runif(n = 20, min = 0, max = 1)) )

ggplot(sumx_sim,
       aes(x = sums)) +
  geom_histogram(color = 'white',
                 binwidth = 0.2)
```

## 4.6

> "Many introductory statistics textbooks say that 'sample sizes equal to or greater than 30 are often considered sufficient for the CLT to hold.' "

Intuitively, this quote gives me the impression that once sample size reaches 30, there is sufficient *n* to assume that our sampling distribution is normal. (I also get the impression that once n = 30, I can treat a continuous variable as independent & identically distributed, which definitely seems a stretch).

```{r}
# original simulation: n = 2247

set.seed(123)
est_prop <- .33
num_sims <- 10000 
svy_size <- 2247  

sims <- tibble(sim_num = 1:num_sims) |> 
  uncount(svy_size)

sims <- sims |> 
  mutate(conservative = rbinom(num_sims*svy_size, 1, est_prop)) |> 
  group_by(sim_num) |> 
  summarize(prop = mean(conservative))

mean(sims$prop)
sd(sims$prop)

lower_bound95 <- quantile(sims$prop, .025)
upper_bound95 <- quantile(sims$prop, .975)
ci95 <- c(lower_bound95, upper_bound95)
ci95

SE <- (sd(sims$prop))/sqrt(2247)
SE
mean(sims$prop) - 1.96*SE

```

```{r}
# n = 30
set.seed(123)
est_prop <- .33
num_sims <- 10000 
svy_size <- 30  

sims <- tibble(sim_num = 1:num_sims) |> 
  uncount(svy_size)

sims <- sims |> 
  mutate(conservative = rbinom(num_sims*svy_size, 1, est_prop)) |> 
  group_by(sim_num) |> 
  summarize(prop = mean(conservative))

mean(sims$prop)
sd(sims$prop)

lower_bound95 <- quantile(sims$prop, .025)
upper_bound95 <- quantile(sims$prop, .975)
ci95 <- c(lower_bound95, upper_bound95)
ci95

SE <- (sd(sims$prop))/sqrt(30)
SE
mean(sims$prop) - 1.96*SE

```

While the means are similar, the standard deviations (and thus standard errors) are quite different, which makes the range of the 95% confidence interval much larger when n = 30. In the case where n=30, the confidence interval is ***not*** within 2 standard errors of the mean. Which leads me to disbelieve the adage that when n=30, conditions for the central limit theorem to hold are met.
