---
title: "Week 5 Homework"
author: "Andrew Joyce"
date: "10-01-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
---

```{r}
#| echo: false
#| results: hide
#| include: false
library(conflicted)
conflicts_prefer(dplyr::filter)
library(tidyverse)
```

## 5.2.1 Exercise

Suppose that `p = 2/3`, and `size = 1`.\
\
Use `dbinom()` to verify the result in Equation 5.1

```{r}
eq1a <- dbinom(1, size = 1, prob = 2/3)
isTRUE(eq1a == 2/3)

eq1b <- dbinom(0, size = 1, prob = 2/3)
isTRUE(eq1b == (1-2/3))

eq1c <- dbinom(5, size = 1, prob = 2/3)
isTRUE(eq1c == 0)


```

Use `pbinom()` to verify the results in Equation 5.2

```{r}
eq2a <- pbinom(-1, size = 1, prob = 2/3)
isTRUE(eq2a == 0)

eq2b <- pbinom(0, size = 1, prob = 2/3)
isTRUE(eq2b == (1 - 2/3))

eq2c <- pbinom(1, size = 1, prob = 2/3)
isTRUE(eq2c == 1)

eq2d <- pbinom(2, size = 1, prob = 2/3)
isTRUE(eq2d == 1)
```

Use `rbinom()` to generate 100 samples from the Bernouilli distribution with `prob = 2/3`. Plot the results in `ggplot` using `geom_bar`.

```{r}
tosses <- tibble(
  heads = (rbinom(100, 1, 2/3)) 
)

ggplot(tosses, aes(x = heads)) + geom_bar(fill = "lightgreen") +
scale_x_continuous(breaks = 0:1)


```

## 5.3.1 Exercise

The sample space ($\Omega$) of four fair coin flips is as follows:

$$
\Omega = 
\begin{Bmatrix}
HHHH \\ \\
HHHT \\ HHTH \\ HTHH \\ THHH \\ \\
HHTT \\ TTHH \\ HTTH \\ THHT \\ HTHT \\ THTH \\ \\
HTTT \\ THTT \\ TTHT \\ TTTH \\ \\
TTTT \\
\end{Bmatrix}  \to
\begin{Bmatrix} 
4 \\ \\
3 \\ 3 \\ 3 \\ 3 \\ \\
2 \\ 2 \\ 2 \\ 2 \\ 2 \\ 2 \\ \\
1 \\ 1 \\ 1 \\ 1 \\ \\
0 \\
\end{Bmatrix}
$$

## 5.3.2 Exercise

"How many possible ways are there to get $X = 0$, $X = 1$, $X = 2$, $X = 3$, and $X = 4$?"

-   There is 1 way to get $X=0$
-   There are 4 ways to get $X = 1$
-   There are 6 ways to get $X = 2$
-   There are 4 ways to get $X = 3$
-   There is 1 way to get $X = 4$.

"What is the probability that $X = 2$?"

-   The probability that $X = 2$ is $\frac{6}{16}$ which is equivalent to $\frac{3}{8}$, $0.375$, or $37.5$%.

"Verify this by using the correct R function to calculate the probability that $X = 2$."

```{r}
dbinom(2, 4, .5)
```

"Use the correct R function to calculate the probability that $X \leq 1$?"

```{r}
dbinom(1, 4, .5) + dbinom(0, 4, .5)
```

## 5.3.3 Exercise

```{r}
# Use rbinom() to generate 1000 samples from the Binomial distribution with size = 5 and prob = 2/3. Plot the results in ggplot using geom_bar.
tosses2 <- tibble(
  heads = (rbinom(1000, 5, 2/3)) 
)

ggplot(tosses2, aes(x = heads)) + geom_bar(fill = "lightblue")
```

```{r}
#Estimate the probability that X is an even number using the mean() function on those 1000 values.

mean(tosses2 == 2 | tosses2 == 4)
```

## 5.3.5 Exercise

"Review 5.3.1 and 5.3.2. What is the probability that $X=2$ when $p = 1/3$ and $n = 4$?"

$$ Pr(X=2) = 6 * \frac{1}{3} \cdot \frac{1}{3} \cdot \frac{2}{3} \cdot \frac{2}{3} = \frac{24}{81} = 0.2963 = 29.63\% $$

\*There are 6 permutations of getting 2 heads when flipping 4 coins, hence multiplying this probability times 6.

"Verify this by using the correct R function to calculate the probability that $X = 2$."

```{r}
dbinom(2, size = 4, prob = 1/3)
```

## 5.3.7 Exercise

**Part 1: Verify equation 5.3 using `dbinom`.**

The equation for the binomial coefficient: $$ {n \choose x} = \frac{n!}{x! (n - x)!} $$

The probabilities of any binomial distribution are:

$$ \Pr(X = x) = {n \choose x} p^x (1-p)^{n-x} $$

I want to verify these, supposing p = 2/3 and size = 15.

I can use `dbinom(X, size, prob)` to perform one calculation.\
I can also hand-calculate using: `choose(size,X)*(prob)^X*(1-prob)^(size-X)`. I

Since `X` will vary, I will create a loop that includes outcomes of \[0, 15\]. That way, I can compare outcomes (0 heads, 1 head, 2 heads....15 heads) using dbinom and the hand calculation to verify they are close (they might not be identical due to rounding). If the two calculations are close, I have successfully verified the equation results.

```{r}
outcomes <- c(1:15)

#dbinom loop
results1 <- numeric(length(outcomes))
for(i in seq_along(outcomes)) {
  results1[i] <- dbinom(i, 15, 2/3)
}

#hand calculation loop
results2 <- numeric(length(outcomes))
for(i in seq_along(outcomes)) {
  results2[i] <- choose(15,i)*((2/3)^i)*((1-2/3)^(15-i))
}

#Comparison of values
tibble(results1, results2)

# sum of each of these results should approximate 1.
sum(results1)
sum(results2)
```

**Part 2: Verify equation 5.4 using `pbinom`.**

```{r}
results3 <- pbinom(q = 0:15, size = 15, prob = 2/3)
results3
```

As q approaches the specified size, the value of the cumulative distribution function approaches 1.

::: callout-warning
I'm not entirely sure how to "verify" an equation.
:::

## 5.4.1 Exercise

```{r}
grid <- tibble(prob = seq(0, 1, by = 0.001))
grid$like <- dbinom(21, 47, grid$prob)
grid <- grid |> 
  mutate(like = dbinom(21, 47, grid$prob),
         clike_raw = cumsum(like),
         clike_normalized = clike_raw/ sum(like))

ggplot(grid,
       aes(x = prob,
           y = clike_normalized)) + geom_line() + 
      geom_vline(xintercept = 0.588, linetype = "dashed", color = "blue") +
      geom_hline(yintercept = 0.975, linetype = "dashed", color = "red" )

```

"What what are the chances that `prob` is equal to or greater than 0.588?"

```{r}
grid |> filter (prob >= 0.588)
```

The chances that `prob` is `>=` 0.588 is 0.975. Or, 97.5%. I graphed this value on the red line on the above graph.
