---
title: "Week 6 Homework"
author: "Andrew Joyce"
date: "10-09-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
---

```{r}
#| echo: false
#| results: hide
#| include: false

library(conflicted)
conflicts_prefer(dplyr::filter)
library(tidyverse)
library(infer)
library(janitor) ## for convenient "table" functions
library(gssr)    ## for access to GSS data

theme_set(theme_light(base_family = "Baskerville")) 
```

## 6.1.2 Exercise

::: callout-note
## Text of Question 6.1.2

You are designing a survey to estimate the proportion of individuals who will vote Democrat. Assuming that respondents are a simple random sample of the voting population, how many people do you need to poll so that the standard error is less than 5 percentage points?
:::

***Method 1: Algebra***

Given the following formula for standard error:

$$se_{p} = \sqrt \frac{p(1-p)}{n}$$ and asssuming $p = 0.53$ & $se_{p}= .05$, solve for $n$.

$$
0.05 = \sqrt \frac{0.53 \times 0.47}{n}
$$ Thus, $n = 99.64$. But, that means that $se_{p} = 0.05$ exactly. The above question asks that the answer is *less than* 5 percentage points. So, I round up: $n = 100$ for $se_{p}$ to be less than 5 percentage points (\< 0.05). (Plus, we can't have 99.64 people, we can only have 99 or 100).

***Method 2: R Code***

I can confirm this with some code. I will calculate the value for $se_{p}$ for values of $n$ ranging from $n=1$ to $n = 1000$. Then, I can see at what value of $n$ is $se_{p} < 0.05$.

```{r}
s_e <- tibble(
  n = c(1:1000),
  std_error = sqrt((0.53*0.47)/n)
)

s_e |> filter(s_e$std_error < 0.05)
```

Thus, when $n=100$, $se_{p} < 0.05$.

## 6.1.5 Exercise

::: callout-note
## Text of 6.1.5 Question, Part 1

What is the probability of observing the "true" value ($p=0.53$) under the null?
:::

To solve:

```{r}
set.seed(321)
S <- 1e4 
poll_size <- 1000 
draws <- rbinom(S, size = poll_size, prob = 0.50)
null <- tibble("prop_est" = draws / poll_size)

nrow(filter(null, prop_est == 0.53)) / nrow(null) * 100

```

The probability that we observe the "true" value ($p=0.53$) under the null is $0.41\%$.

::: callout-note
## Text of 6.1.5 Question, Part 2

What is the probability of observing `prop_hat` under the null? Is this statistically significant if the confidence level ($\alpha$) is set to 0.05?
:::

```{r}
set.seed(321)
one_dataset <- rbinom(poll_size, size = 1, prob = 0.53)
prop_hat <- mean(one_dataset)

nrow(filter(null, prop_est == prop_hat)) / nrow(null) * 100 
```

The probability of observing `prop_hat` under the null is 1.75%. If $\alpha$ is set to 0.05, that is statistically significant, since `prop_hat` \< $\alpha$.

## 6.1.7 Exercise

::: callout-note
## Text of 6.1.7 Exercise Question

The formula for calculating the standard error of a difference in proportions is given by:

$$
\sigma_{\hat p_1 - \hat p_2} = \sqrt{\frac{p_1 (1 - p_1)}{n_1} + \frac{p_2 (1 - p_2)}{n_2}}
$$

Verify that this standard error corresponds to `sd(theta_distribution)`.
:::

I can verify that the above equation is equal to `sd(theta_distribution)` by computing the value of $\sigma_{\hat p_1 - \hat p_2}$ mathematically.

```{r}
# first, let's generate the same numbers as in the example
p1 <- 0.5
n1 <- 120
p2 <- 0.6
n2 <- 90

S <- 1e5
draws1 <- rbinom(S, size = n1, prob = p1) 
proportions1 <- draws1 / n1 
draws2 <- rbinom(S, size = n2, prob = p2)
proportions2 <- draws2 / n2
theta_distribution <- proportions1 - proportions2

```

Now that I have the example numbers in memory, I can verify:

```{r}
sqrt( ((p1 * (1- p1))/n1) + ((p2 * (1- p2))/n2) )
sd(theta_distribution)
```

I can also graph the above equation on top of the histogram for `sd(theta_distribution)`.

```{r}
tibble(theta_distribution) |> 
  ggplot(aes(theta_distribution)) +
  geom_histogram(
    color = "white", boundary = 0.5, binwidth = 0.025,
    mapping = aes(y = after_stat(density))
  ) +
  labs(title = "Sampling Distribution of Differences in Proportions",
       subtitle= "P1 = 0.5, P2 = 0.6, N1 = 120, N2 = 90") +
  geom_function(fun = \(x) dnorm(x, mean = mean(theta_distribution), sd = sd(theta_distribution)), color = "blue", linewidth = 1.05)
```

## 6.1.8 Exercise

::: callout-note
## Text of 6.1.8 Exercise Question

*Comparison of proportions.*

A randomized experiment is performed within a survey. 1000 people are contacted. Half the people contacted are promised a \$5 incentive to participate, and half are not promised an incentive. The result is a 50% response rate among the treated group and 40% response rate among the control group. Give an estimate and standard error of the difference in proportions.
:::

```{r}
p1 <- 0.5
n1 <- 500
p2 <- 0.4
n2 <- 500

S <- 1e4
draws1 <- rbinom(S, size = n1, prob = p1) 
proportions1 <- draws1 / n1 
draws2 <- rbinom(S, size = n2, prob = p2)
proportions2 <- draws2 / n2
theta_distribution <- proportions1 - proportions2

# estimate of diff of proportions
mean(theta_distribution)

# se of the difference in proportions
sd(theta_distribution)
```

The estimate for the difference in proportions is 9.95%. The standard error of the difference in proportions is 0.0315.

## GSS Setup

```{r}
gss18 <- gss_get_yr(2018) 

d <- gss18 |> 
  select(sex, attend, polviews) |> 
  haven::zap_missing() |> 
  mutate(sex = as_factor(sex)) |> 
  haven::zap_labels() |> 
  drop_na()

glimpse(d)
```

## 6.2.1 Exercise

The variable `attend` asks respondents "How often do you attend religious services?" The answers are coded as follows:

-   0: Never
-   1: Less than once a year
-   2: About once or twice a year
-   3: Several times a year
-   4: About once a month
-   5: 2-3 times a month
-   6: Nearly every week
-   7: Every week
-   8: Several times a week

The variable `polviews` asks respondents to place themselves on a scale of political views from 1 to 7, with categories being:

-   1: Extremely liberal
-   2: Liberal
-   3: Slightly liberal
-   4: Moderate, middle of the road
-   5: Slightly conservative
-   6: Conservative
-   7: Extremely conservative

## 6.2.2 Exercise

::: callout-note
## Text of 6.2.2 Exercise Question

Repeat what we did in class with Steve, but compare the `weekly` variable to a new variable call `conservative`.
:::

```{r}
d <- gss18 |> 
  mutate(female = if_else(sex == 2, 1L, 0L),
         weekly = if_else(attend >= 7, 1L, 0L),
         conservative = if_else(polviews >= 5, 1L, 0L)) |> 
  select(female, weekly, conservative) |> 
  drop_na() |> 
  mutate(sex = if_else(female == 1, "Female", "Male"))

glimpse(d$conservative)

```

## 6.2.3 Exercise

::: callout-note
## Question

Is the difference in proportions between `conservative` and `weekly` statistically significant?
:::

```{r}
obs_diff <- mean(d$conservative) - mean(d$weekly)
d <- d |> mutate(attendance = if_else(weekly == 1, "Weekly", "LessThanWeekly"))

boot_dist <- d |> 
  specify(conservative ~ attendance) |> 
  generate(reps = 1000,
           type = "bootstrap") |> 
  calculate(stat = "diff in means", 
            order = c("Weekly", "LessThanWeekly")) 

ci <- boot_dist |> 
  get_confidence_interval(level = .95)
ci

boot_dist |> 
  visualize() +
  shade_ci(ci) 



```

The 95% confidence interval in the difference of proportions is between 14.7% and 24.7%. Thus, the difference in proportions between `conservative` and `weekly` is significant.

## 6.2.4 Exercise

The grid "counts" shows the cross-tabulation for the 9 total categories of `attend` (0 through 8) against the 7 total categories of `polviews` (1 through 7). Each box in the grid shows the number of respondents in each category. For example, there are 16 respondents who report 0 for attend (never) and 7 for polviews (extremely conservative), which is the cell in the top left corner.

The grid "row percentages" calculates for each cell the proportion of respondents relative to total number of respondents in that row. Here, the rows are the categories of responses to `polviews`. There are 122 total (69 + 5 + 9 + 7 + 3 + 9 + 3 + 12 + 5) respondents who selected "1: Extremely liberal" on their response to `polviews`. Each cell calculates the percentage of respondents based on their reported attendance. For example, 69 out of the total 122 respondents for 1 on polviews also selected 0 on attendance. 69/122 = 56.6%. Then, of the 122 respondents who report 1 on polviews, 5 of these selected 1 on attendance. 5/122 = 4.0%. In short, a row percentage shows the proportion of respondents in that row (here, a response to the polviews question) in a particular category of the column (here, a response to the attendance question.)

The grid "column percentages" does the reverse. Each cell is the percent of respondents within that column that chose a particular row response. Here, the columns are the categories of responses to `attend`. There are 668 total (16 + 74 + 66 + 248 + 77 + 128 + 69) respondents who selected "0" on their response to attend (0: never). Each cell in that column is the proportion of those who selected 0 on their attend response for each category of `polviews`. For example, 16 out of 668 total who selected 0 on attend also selected 7 on polviews (extremely conservative). 16/668 = 2.4%.

In sum: row percentages are the % that each each cell represents of the *row total*.\
Column percentages are the % that each cell represents of the *column total.*

This can get confusing. Note that the sum of all cells is far greater than 100. Rather, the sum of all the percentages in a row is 100, and the sum of all percentages in a column is 100. This is because these percentages are based on row/column totals, not the total sample.

The % in a cell is *not* the total number of cases in the sample that have both characteristics of row and column values. For a row percentage, the % in a cell is the proprotion of cases *from all cases **in that row*** with the particular column characteristic. For a column percentage, the % in a cell is the proportion of cases *from all cases **in that column*** with the particular row characteristic. Tricky!
