---
title: "Week 7 Homework"
author: "Andrew Joyce"
date: "10-16-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
bibliography: references.bib
---

```{r}
#| echo: false
#| results: hide
#| include: false

library(conflicted)
conflicts_prefer(dplyr::filter)
library(tidyverse)
library(infer)
library(janitor) ## for convenient "table" functions
library(gssr)    ## for access to GSS data

theme_set(theme_light(base_family = "Baskerville")) 
```

## 7.1.1. Exercise

```{r}
#| echo: false
#| results: hide
#| include: false

library(tidyverse)
library(gssr)
gss18v1 <- gss_get_yr(2018) 

d1 <- gss18v1 |> 
  select(attend, polviews) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(weekly = if_else(attend >= 7, 1L, 0L),
         conservative = if_else(polviews >= 5, 1L, 0L)) |> 
  drop_na() 
```

```{r}
tabyl(d1, weekly, conservative) |> adorn_totals(c("row", "col") )

```

```{r}
tabyl(d1, weekly) ## for answers #1 and #2
tabyl(d1, conservative) # for answer number 8
```

1.  $\Pr(W)$ = 0.22
2.  $\Pr(W^C)$ = 0.77

```{r}
# to answer #3 and #5
tabyl(d1, weekly, conservative) |> 
  adorn_percentages() |> 
  adorn_pct_formatting(digits = 0)

# to answer #4
tabyl(d1, weekly, conservative) |> 
  adorn_percentages(denominator = "col") |> 
  adorn_pct_formatting(digits = 0)

```

3.  $\Pr(C \mid W)$ = 0.48
4.  $\Pr(W \mid C)$ = 0.33
5.  $\Pr(C \mid W^\complement)$ = 0.28

```{r}
# for answers to #6 and #7
tabyl(d1, weekly, conservative) |> 
  adorn_percentages(denominator = "all") |>  
  adorn_pct_formatting(digits = 0)
```

6.  $\Pr(W \text{ and } C)$ = 0.11
7.  $\Pr(W^\complement \text{ and } C^\complement)$ = 0.56
8.  If we assume $W$ and $C$ are independent, what *would* be their joint probability?

Events $W$ and $C$ are Independent if and only if the joint probability is equal to the product of the marginal probabilities. That is, $\Pr(W \text{ and } C) = \Pr(W)Pr(C)$.

Since $\Pr(W)$ = 0.22 and $\Pr(C)$ = 0.33, we should expect the joint probability to be $0.22 * 0.33$, or 0.0726

Thus, the **expected joint probability assuming independence = 0.0726, or 7.26%.**

## 7.1.2 Exercise

@diprete2013 find that women outnumber men in terms of bachelor's degree completion. So, I am going to compare sex ($F$ = female, $F^\complement$=male) and completion of a bachelor's degree or higher ( $D$ ).

```{r}
# data prep
library(gssr)
gss18v2 <- gss_get_yr(2018) |> 
  haven::zap_labels()

d2 <- gss18v2 |> 
  select(sex, degree) |> 
  drop_na() |> 
  mutate(female = if_else(sex == 2, 1L, 0L),                       
         sex = if_else(female == 1, "Female", "Male"),             
         degree = if_else(degree >= 3, 1L, 0L),                  
         degree_chr = if_else(degree == 1, "Bachelor's or Higher", "Less than Bachelor's")) |> 
  relocate(sex, female, degree_chr, degree)
```

```{r}
# contingency table
tabyl(d2, sex, degree_chr) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages(denominator = "all") |> 
  adorn_pct_formatting(digits = 0) |> 
  adorn_ns() 
 
```

I will show relevant code below, but I will input the answers to the questions in the table here.

+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| Notation                                                 | Translation                                                                                             | Probability                                          |
+==========================================================+=========================================================================================================+======================================================+
| 1.  $\Pr(D)$                                             | Probability of completing a bachelor's degree or higher                                                 | 0.30 or 30%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 2.  $\Pr(D^\complement)$                                 | Probability of not completing a bachelor's degree or higher                                             | 0.70 or 70%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 3.  $\Pr(F \mid D)$                                      | Probability of being female given completion a college degree (note that this is a nonsense prediction) | 0.56 or 56%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 4.  $\Pr(D \mid F)$                                      | Probability of completing a college degree given respondent is female                                   | 0.31 or 31%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 5.  $\Pr(D \mid F^\complement)$                          | Probability of completing a college degree given respondent is male (literally, not female)             | 0.30 or 30%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 6.  $\Pr(D \text{ and } F)$                              | Joint probability of completing a college degree and female                                             | 0.17 or 17%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 7.  $\Pr(D^\complement \text{ and } F^\complement)$      | Joint probability of not completing a college degree and male                                           | 0.31 or 31%                                          |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+
| 8.  Expected $\Pr(D \text{ and } F)$ given independence. | Assuming sex and degree completion are independent, what would we expect the joint probability to be?   | Expected $\Pr(D \text{ and } F)$ given independence: |
|                                                          |                                                                                                         |                                                      |
|                                                          |                                                                                                         | $Pr(D) * Pr(F)$ = $0.30 * 0.55$ = $0.165$            |
|                                                          |                                                                                                         |                                                      |
|                                                          |                                                                                                         | or **16.5% assuming independence**                   |
+----------------------------------------------------------+---------------------------------------------------------------------------------------------------------+------------------------------------------------------+

```{r}
# for 1 & 2 and #8
tabyl(d2, degree)
tabyl(d2, female)

# to answer #3  (Pr(F | D))
tabyl(d2, degree, female) |> 
  adorn_percentages() |> 
  adorn_pct_formatting(digits = 0)

# to answer #4 and 5 (Pr(D | F) and Pr (D | F^C))
tabyl(d2, degree, female) |> 
  adorn_percentages(denominator = "col") |> 
  adorn_pct_formatting(digits = 0)

# to answer #7 (Pr(D & F))
tabyl(d2, degree, female) |> 
  adorn_percentages(denominator = "all") |> 
  adorn_pct_formatting(digits = 0)
```

## 7.2.1 Exercise

Comparing the probabilities of $\Pr(W \mid C)$ and $\Pr(W \mid C^\complement)$

-   $\Pr(W \mid C)$ = 0.33

-   $\Pr(W \mid C^\complement)$ = 0.17

+-----------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Method                      | Result                                                                                                                                                        |
+=============================+===============================================================================================================================================================+
| Difference in Probabilities | 0.16 or 16% (probability of weekly attendance for conservatives is 16% greater than the probability of weekly attendance for those who are not conservatives) |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Relative Risk Ratio         | 1.94 (probability of weekly attendance is 1.94 times greater for conservatives than for those who are not conservatives)                                      |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Odds Ratio                  | 2.35 (the odds of weekly attendance are 2.35 times greater for conservatives than for those who are not conservatives)                                        |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Log Odds ratio              | 0.85 (the log-odds of weekly attendance are 0.85 units higher for conservatives than for those who are not conservative)                                      |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}
# diff in probabilities
0.33 - 0.17

# relative risk ratio
0.33/0.17

# odds ratio
(1243 * 240)/(257 * 495)

# log-odds ratio
log((1243 * 240)/(257 * 495))
```

## 7.2.2 Exercise

1.  Draw a sampling distribution for the relative risk ratio `(stat = "ratio of props")`

```{r}
boot1 <- d1 |> 
  mutate(conservative = if_else(conservative == 1L, "conservative", "other")) |> 
  mutate(weekly = if_else(weekly == 1L, "weekly", "other")) |> 
  specify(response = weekly, explanatory = conservative, success = "weekly") |> 
  generate(reps = 1e4, type = "bootstrap") |> 
  calculate(stat = "ratio of props", order = c("conservative", "other"))
  
ci <- get_confidence_interval(boot1)

boot1 |> 
  visualize() +
  shade_ci(ci) +
  geom_vline(xintercept = 1.94, linetype = "dashed") # this is the value I hand-calculated in 7.2.1

```

2.  Draw a sampling distribution for the odds-ratio

```{r}
boot1 <- d1 |> 
  mutate(conservative = if_else(conservative == 1L, "conservative", "other")) |> 
  mutate(weekly = if_else(weekly == 1L, "weekly", "other")) |> 
  specify(response = weekly, explanatory = conservative, success = "weekly") |> 
  generate(reps = 1e4, type = "bootstrap") |> 
  calculate(stat = "odds ratio", order = c("conservative", "other"))
  
ci <- get_confidence_interval(boot1)

boot1 |> 
  visualize() +
  shade_ci(ci) +
  geom_vline(xintercept = 2.35, linetype = "dashed") # this is the value I hand-calculated in 7.2.1
```

## 7.2.3 Exercise

Comparing the probabilities of $\Pr(D \mid F)$ and $\Pr(D \mid F^\complement)$

-   $\Pr(D \mid F)$ = 0.31 (Probability of college degree completion, given female)

-   $\Pr(D \mid F^\complement)$ = 0.30 (Probability of college degree completion, given male)

+-----------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| Method                      | Result                                                                                                                              |
+=============================+=====================================================================================================================================+
| Difference in Probabilities | 0.01 or 1% (probability of college degree completion for women is 1% greater than probability of college degree completion for men) |
+-----------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| Relative Risk Ratio         | 1.03 (probability of weekly attendance is 1.03 times greater for women than for men)                                                |
+-----------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| Odds Ratio                  | 1.02 (the odds of degree completion are 1.02 times greater for women than for men)                                                  |
+-----------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| Log Odds ratio              | 0.02 (the log-odds of degree completion are 0.02 units greater for women than for men)                                              |
+-----------------------------+-------------------------------------------------------------------------------------------------------------------------------------+

*Note: I suspect these differences are low because the GSS samples adults of all ages (18-89). The findings from @diprete2013 are of newer cohorts.*

```{r}
# diff in probabilities
0.31 - 0.30

# relative risk ratio
0.31/0.30

# odds ratio
(396 * 736)/(900 * 316)

# log-odds ratio
log((396 * 736)/(900 * 316))
```

## 7.2.4 Exercise

1.  Draw a sampling distribution for the relative risk ratio `(stat = "ratio of props")`

```{r}
boot2 <- d2 |> 
  mutate(female = if_else(female == 1L, "women", "men")) |> 
  mutate(degree = if_else(degree == 1L, "degree", "other")) |> 
  specify(response = degree, explanatory = female, success = "degree") |> 
  generate(reps = 1e4, type = "bootstrap") |> 
  calculate(stat = "ratio of props", order = c("women", "men"))
  
ci <- get_confidence_interval(boot2)

boot2 |> 
  visualize() +
  shade_ci(ci) +
  geom_vline(xintercept = 1.03, linetype = "dashed") # this is the value I hand-calculated in 7.2.3
```

2.  Draw a sampling distribution for the odds ratio.

```{r}
boot2 <- d2 |> 
  mutate(female = if_else(female == 1L, "women", "men")) |> 
  mutate(degree = if_else(degree == 1L, "degree", "other")) |> 
  specify(response = degree, explanatory = female, success = "degree") |> 
  generate(reps = 1e4, type = "bootstrap") |> 
  calculate(stat = "odds ratio", order = c("women", "men"))
  
ci <- get_confidence_interval(boot2)

boot2 |> 
  visualize() +
  shade_ci(ci) +
  geom_vline(xintercept = 1.02, linetype = "dashed") # this is the value I hand-calculated in 7.2.3
```
