---
title: "Week 8 Homework"
author: "Andrew Joyce"
date: "10-21-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
toc: true
---

## Two Models

***Linear Probability Model***

-   General form: $\Pr(Y = 1 \mid X = x) = \alpha +\beta x$
-   Simplified form: $p = \alpha + \beta x$
-   $\alpha = \Pr(Y = 1 \mid X = 0)$
-   $\beta = \Pr(Y = 1 \mid X = 1) - \Pr(Y = 1 \mid X = 0)$

Here, $\alpha$ is a probability and $\beta$ is a difference in probabilities.

***Logistic Regression Model***

-   General form: $\log (\frac{p}{1-p}) = \alpha + \beta x$
-   $\alpha = \log \Big( \frac{\Pr(Y = 1 \mid X = 0)}{\Pr(Y = 0 \mid X = 0)}  \Big)$
-   $\alpha + \beta =  \log \Big( \frac{\Pr(Y = 1 \mid X = 1)}{\Pr(Y = 0 \mid X = 1)}  \Big)$

Here, $\alpha$ is a log-dd and $\beta$ is a difference in log-odds (a log-odds ratio).

## Setup: attend, degree, sex, and polviews

```{r}
#| echo: false
#| results: hide
#| include: false

library(conflicted)
conflicts_prefer(dplyr::filter)
library(tidyverse)
library(infer)
library(janitor) ## for convenient "table" functions
library(gssr)    ## for access to GSS data

theme_set(theme_light(base_family = "Baskerville")) 
```

```{r}
gss18 <- gss_get_yr(2018) 

d <- gss18 |> 
  select(attend, polviews, sex, degree) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(weekly = if_else(attend >= 7, 1L, 0L),
         conservative = if_else(polviews >= 5, 1L, 0L),
         female = if_else(sex == 2, 1L, 0L),
         college = if_else(degree >= 3, 1L, 0L)) |> 
  drop_na() 
```

```{r}
# glm(formula = weekly ~ college, family = "binomial", data = d)

## general form
# glm(formula = y ~ x, family = "binomial", data = d)
```

## Model 1: Weekly & College

For variable "weekly":

-   1: Attends religious services weekly or more
-   0: Attends religious services less than weekly

For variable "college":

-   1: Has completed a bachelor's degree or higher
-   0: Has not completed a bachelor's degree

Y = weekly, X = college

***Linear Probability model:***

$\alpha = 0.205$

$\beta = 0.061$

$p = 0.205 + 0.061x$

The probability of those with a college degree attending religious services weekly is 6.1% points greater than the probability of those without a college degree attending religious services weekly.

***Logistic Regression model:***

$\alpha = -1.352$

$\beta = 0.304$

$\text{logit} (p) = 0.304x - 1.352$

Those with a college degree have 0.304 higher logged odds of attending religious services weekly compared to those who do not have a college degree.

```{r}
tabyl(d, weekly, college) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
316/1538

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(181/679) - (316/1538)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((316/1538)/(1222/1538))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((181/697)/(516/697)) - (log((316/1538)/(1222/1538)))
```

## Model 2: Weekly & Female

For variable "weekly":

-   1: Attends religious services weekly or more
-   0: Attends religious services less than weekly

For variable "female":

-   1: Respondent is a woman
-   0: Respondent is a man

Y = weekly, X = female

***Linear Probability model:***

$\alpha = 0.203$

$\beta = 0.035$

$p = 0.203 + 0.035x$

The probability of women attending religious services weekly is 3.5% points greater than the probability of men attending religious services weekly.

***Logistic Regression model:***

$\alpha = -1.367$

$\beta = 0.204$

$\text{logit} (p) = 0.204x - 1.367$

Women have 0.204 higher logged odds of attending religious services weekly compared to men.

```{r}
tabyl(d, weekly, female) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
205/1008

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(292/1226) - (205/1008)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((205/1009)/(804/1009))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((292/1226)/(934/1226)) - log((205/1009)/(804/1009))
```

## Model 3: Weekly & Conservative

For variable "weekly":

-   1: Attends religious services weekly or more
-   0: Attends religious services less than weekly

For variable "conservative":

-   1: Respondent describes political views as conservative
-   0: Respondent describes political views as something other than conservative

Y = weekly, X = conservative

***Linear Probability model:***

$\alpha = 0.171$

$\beta = 0.155$

$p = 0.171 + 0.155x$

The probability of conservatives attending religious services weekly is 15.5% points greater than the probability of those who are not conservative attending religious services weekly.

***Logistic Regression model:***

$\alpha = -1.576$

$\beta = 0.852$

$\text{logit} (p) = 0.852x - 1.576$

Conservatives have 0.852 higher logged odds of attending religious services weekly compared to those who are not conservative.

```{r}
tabyl(d, weekly, conservative) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
257/1500

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(240/735) - (257/1500)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((257/1500)/(1243/1500))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((240/735)/(495/735)) - log((257/1500)/(1243/1500))
```

## Model 4: College & Female

For variable "college":

-   1: Has completed a bachelor's degree or higher
-   0: Has not completed a bachelor's degree

For variable "female":

-   1: Respondent is a woman
-   0: Respondent is a man

Y = college, X = female

***Linear Probability model:***

$\alpha = 0.310$

$\beta = 0.003$

$p = 0.310 + 0.003x$

The probability of women completing a college degree is 0.3% points greater than the probability of men completing a college degree

***Logistic Regression model:***

$\alpha = -0.799$

$\beta = 0.014$

$\text{logit} (p) = 0.014x - 0.799$

Women have 0.014 higher logged odds of completing a college degree compared to men.

```{r}
tabyl(d, college, female) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
313/1009

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(384/1226) - (313/1009)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((313/1009)/(696/1009))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((384/1226)/(842/1226)) - log((313/1009)/(696/1009))
```

## Model 5: College & Weekly

For variable "college":

-   1: Has completed a bachelor's degree or higher
-   0: Has not completed a bachelor's degree

For variable "weekly":

-   1: Attends religious services weekly or more
-   0: Attends religious services less than weekly

Y = college, X = weekly

***Linear Probability model:***

$\alpha = 0.297$

$\beta = 0.182$

$p = 0.297 + 0.182x$

The probability of weekly attenders completing a college degree is 18.2% points greater than the probability of those who do not attend weekly completing a college degree.

***Logistic Regression model:***

$\alpha = -0.862$

$\beta = 0.305$

$\text{logit} (p) = 0.305x - 0.862$

Those who attend religious services weekly have 0.305 higher logged odds of completing a college degree compared to those who attend religious services at lower rates.

```{r}
tabyl(d, college, weekly) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
516/1738

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(181/497) - (316/1738)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((516/1738)/(1222/1738))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((181/497)/(316/497)) - log((516/1738)/(1222/1738))
```

## Model 6: College & Conservative

For variable "college":

-   1: Has completed a bachelor's degree or higher
-   0: Has not completed a bachelor's degree

For variable "conservative":

-   1: Respondent describes political views as conservative
-   0: Respondent describes political views as something other than conservative

Y = college, X = weekly

***Linear Probability model:***

$\alpha = 0.318$

$\beta = -0.019$

$p = 0.318 - 0.019x$

The probability of conservatives completing a college degree is 1.9% points less than the probability of those who are not conservative completing a college degree.

***Logistic Regression model:***

$\alpha = -0.763$

$\beta = -0.088$

$\text{logit} (p) = -0.088x - 0.763$

Conservatives have 0.088 lower logged odds of completing a college degree compared to those who are not conservative.

```{r}
tabyl(d, college, conservative) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
477/1500

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(220/735) - (477/1500)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((477/1500)/(1023/1500))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((220/735)/(515/735)) - log((477/1500)/(1023/1500))
```

## Model 7: Conservative & College

For variable "conservative":

-   1: Respondent describes political views as conservative
-   0: Respondent describes political views as something other than conservative

For variable "college":

-   1: Has completed a bachelor's degree or higher
-   0: Has not completed a bachelor's degree

Y = conservative, X = college

***Linear Probability model:***

$\alpha = 0.335$

$\beta = -0.019$

$p = 0.335 - 0.019x$

The probability of those who have a college degree identifying as conservative is 1.9% points lower than the probability for those who have not completed a college degree.

***Logistic Regression model:***

$\alpha = -0.686$

$\beta = -0.088$

$\text{logit} (p) = -0.088x - 0.686$

Those who have a college degree have 0.088 lower logged odds of identifying as a college degree compared to those who have not completed a college degree.

```{r}
tabyl(d, conservative, college) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
515/1538

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(220/697) - (515/1538)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((515/1538)/(1023/1538))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((220/697)/(477/697)) - log((515/1538)/(1023/1538))
```

## Model 8: Conservative & Female

For variable "conservative":

-   1: Respondent describes political views as conservative
-   0: Respondent describes political views as something other than conservative

For variable "female":

-   1: Respondent is a woman
-   0: Respondent is a man

Y = conservative, X = college

***Linear Probability model:***

$\alpha = 0.347$

$\beta = -0.033$

$p = 0.347 - 0.033x$

The probability of women identifying as conservative is 3.3% points lower than the probability for men identifying as conservative.

***Logistic Regression model:***

$\alpha = -0.633$

$\beta = -0.149$

$\text{logit} (p) = -0.149x - 0.633$

Women have 0.149 lower logged odds of identifying as conservative compared to men.

```{r}
tabyl(d, conservative, female) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
350/1009

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(385/1226) - (350/1009)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((350/1009)/(659/1009))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((385/1226)/(841/1226)) - log((350/1009)/(659/1009))

# verify logistic answers using a regression model
glm(formula = conservative ~ female, family = "binomial", data = d)
```
