---
title: "Week 9 Homework"
author: "Andrew Joyce"
date: "10-28-2023"
format: html
editor: visual
mainfont: "Baskerville"
embed-resources: true
toc: true
---

## Setup

```{r}
#| results: hide
#| error: false
#| warning: false
#| message: false

library(conflicted)
conflicts_prefer(dplyr::filter)

library(tidyverse)
library(gssr)
library(janitor)
theme_set(theme_light(base_family = "Optima"))

gss18 <- gss_get_yr(2018) 

d <- gss18 |> 
  select(attend, polviews, cappun) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(
    weekly = if_else(attend >= 7, 1L, 0L),
    conservative = if_else(polviews >= 5, 1L, 0L),
    cappun = if_else(cappun == 1, 1L, 0L)
    ) |> 
  drop_na() |> 
  select(cappun, weekly, conservative) 
```

## 9.1.1 Exercise (2 x 2)

```{r}
tabyl(d, cappun, weekly) |> adorn_totals(c("row", "col"))

# GLM alpha: Pr(Y = 1 | X = 0)
1050/1637

# GLM Beta: Pr(Y = 1 | X = 1) - Pr (Y = 1 | X = 0)
(277/465) - (1050/1637)

# Logistic alpha: log( (pr(y=1 | x=0))/ pr(y = 0 | x = 0))
log((1050/1637)/(587/1637))

# Logistic beta: log ( (pr(y = 1 | x = 1)) / pr(y = 0 | x = 1)) - alpha
log((277/465)/(188/465)) - log((1050/1637)/(587/1637))

# verify calculation for logistic model
glm(formula = cappun ~ weekly, family = "binomial", data = d)
```

$X$ = Weekly attendance and $Y$ = favor capital punishment

***Linear Probability model:***

$\alpha = 0.641$

$\beta = -0.046$

$p = 0.641 - 0.0461x$

***Logistic Model***

$\alpha = 0.582$

$\beta = -0.194$

$\text{logit} (p) = 0.582 - 0.194x$

## 9.1.2 Exercise (2 x 2 x 2)

```{r}
crosstab2 <- table(cp = d$cappun, weekly = d$weekly, conservative = d$conservative)

as_tibble(crosstab2) |> 
  mutate(cp = if_else(cp == "1", "favors", "opposes")) |> 
  pivot_wider(names_from = cp, values_from = n)

# logistic alpha: x1 = 0 and x2 = 0
log(682/485)

# logistic beta1: x1 = 1 and x2 = 0
log(123/115) - log(682/485)

# logistic beta2: x1 = 0 and x2 = 1
log(368/102) - log(682/485)

# logistic beta3: x1 = 1 and x2 = 1
log(154/73) - (log(123/115) - log(682/485)) - (log(368/102) - log(682/485)) - log(682/485)

# verify
glm(formula = cappun ~ weekly + conservative + weekly*conservative, family = "binomial", data = d)

# LPM alpha: Pr(y = 1 | x1=0 & x2 = 0)
682/(485 + 682)

# LPM beta1: Pr(y = 1 | x1=1 & x2 = 0) - alpha
(123/(123 + 115)) - 682/(485 + 682)

# LPM beta2: Pr(y = 1 | x1=0 & x2 = 1) - alpha
(368/(368+102)) - 682/(485 + 682)

# LPM beta3: Pr(y = 1 | x1=1 & x2 = 1) - beta1 - beta2 - alpha
(154/(154+73)) - ((123/(123 + 115)) - 682/(485 + 682)) - ((368/(368+102)) - 682/(485 + 682)) - (682/(485 + 682))


```

Given `cappun` ( $Y$ ), `weekly` ( $X_1$ ), and `conservative` ( $X_2$ ):

***Logistic Model***

$\alpha = 0.341$

$\beta_1 = -0.274$

$\beta_2 = 0.942$

$\beta_3 = -0.263$

$\text{logit} (p) = 0.341 - 0.274x_1 + 0.942x_2 - 0.263x_1x_2$

***Linear Probability Model***

$\alpha = 0.584$

$\beta_1 = -0.068$

$\beta_2 = 0.199$

$\beta_3 = -0.037$

$\text{Pr(favoring capital punishment)} = 0.584 - 0.068x_1 + 0.199x_2 - 0.037x_1x_2$

## 9.2 Exercise (2 x 2 x 2 with a second set of variables)

`college` ( $Y$ ), `macollege` ( $X_1$ ), and `pacollege` ( $X_2$ )

-   Where `college` = 1 if respondent has completed a bachelor's degree or higher, and `college` = 0 otherwise
-   `macollege` = 1 if respondent's mother has completed a bachelor's degree or higher, and `macollege` = 0 otherwise
-   `pacollege` = 1 if respondent's father has completed a bachelor's degree or higher, and `pacollege` = 0 otherwise

```{r}
d2 <- gss18 |> 
  select(degree, madeg, padeg) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(
    college = if_else(degree >= 3, 1L, 0L),
    macollege = if_else(madeg >= 3, 1L, 0L),
    pacollege = if_else(padeg >= 3, 1L, 0L)
    ) |> 
  drop_na() |> 
  select(college, macollege, pacollege) 

crosstab3 <- table(cp = d2$college, macollege = d2$macollege, pacollege = d2$pacollege)

as_tibble(crosstab3) |> 
  mutate(cp = if_else(cp == "1", "BA +", "< BA")) |> 
  pivot_wider(names_from = cp, values_from = n)

# logistic alpha: x1 = 0 and x2 = 0
log(303/880)

# logistic beta1: x1 = 1 and x2 = 0
log(63/53) - log(303/880)

# logistic beta2: x1 = 0 and x2 = 1
log(83/69) - log(303/880)

# logistic beta3: x1 = 1 and x2 = 1
log(141/54) - (log(63/53) - log(303/880)) - (log(83/69) - log(303/880)) - log(303/880)

# verify
glm(formula = college ~ macollege + pacollege + macollege*pacollege, family = "binomial", data = d2)

# LPM alpha: Pr(y = 1 | x1=0 & x2 = 0)
303/1183

# LPM beta1: Pr(y = 1 | x1=1 & x2 = 0) - alpha
(63/116) - (303/1183)

# LPM beta2: Pr(y = 1 | x1=0 & x2 = 1) - alpha
(83/152) - (303/1183)

# LPM beta3: Pr(y = 1 | x1=1 & x2 = 1) - beta1 - beta2 - alpha
(141/195) - ((63/116) - (303/1183)) - ((83/152) - (303/1183)) - (303/1183)
```

***Logistic Model***

$\alpha = -1.066$

$\beta_1 = 1.239$

$\beta_2 = 1.251$

$\beta_3 = -0.464$

$\text{logit} (p) = -1.066 + 1.239x_1 + 1.251x_2 - 0.464x_1x_2$

***Linear Probability Model***

$\alpha = 0.256$

$\beta_1 =  0.287$

$\beta_2 = 0.290$

$\beta_3 = -0.110$

$\text{Pr(completing bachelor's degree)} = 0.256 + 0.287x_1 + 0.290x_2 - 0.110x_1x_2$

## 9.3 Exercise (2 x 2 x 2 with a third set of variables)

`edspend` ( $Y$ ), `weekly` ( $X_1$ ), and `conservative` ( $X_2$ )

Where `edspend` = 1 if respondent thinks we spend "too little" on education, and `edspend` = 0 if respondent thinks we spend "about right" or "too much" on education (based on `nateduc`)

```{r}
d3 <- gss18 |> 
 select(nateduc, attend, polviews) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(
    edspend = if_else(nateduc == 1, 1L, 0L),
    weekly = if_else(attend >= 7, 1L, 0L),
    conservative = if_else(polviews >= 5, 1L, 0L),
    ) |> 
  drop_na() |> 
  select(edspend, weekly, conservative) 

crosstab4 <- table(cp = d3$edspend, weekly = d3$weekly, conservative = d3$conservative)

as_tibble(crosstab4) |> 
  mutate(cp = if_else(cp == "1", "Too Little (Pro Spending)", "Too Much/About Right")) |> 
  pivot_wider(names_from = cp, values_from = n)

# logistic alpha: x1 = 0 and x2 = 0
log(492/129)

# logistic beta1: x1 = 1 and x2 = 0
log(103/21) - log(492/129)

# logistic beta2: x1 = 0 and x2 = 1
log(179/72) - log(492/129)

# logistic beta3: x1 = 1 and x2 = 1
log(76/50) - (log(103/21) - log(492/129)) - (log(179/72) - log(492/129)) - log(492/129)

# verify
glm(formula = edspend ~ weekly + conservative + weekly*conservative, family = "binomial", data = d3)

# LPM alpha: Pr(y = 1 | x1=0 & x2 = 0)
492/621

# LPM beta1: Pr(y = 1 | x1=1 & x2 = 0) - alpha
(103/124) - (492/621)

# LPM beta2: Pr(y = 1 | x1=0 & x2 = 1) - alpha
(179/251) - (492/621)

# LPM beta3: Pr(y = 1 | x1=1 & x2 = 1) - beta1 - beta2 - alpha
(76/126) - ((103/124) - (492/621)) - ((179/251) - (492/621)) - (492/621)
```

***Logistic Model***

$\alpha = 1.339$

$\beta_1 = 0.252$

$\beta_2 = -0.428$

$\beta_3 = -0.744$

$\text{logit} (p) = 1.339 + 0.252x_1 - 0.428x_2 - 0.744x_1x_2$

***Linear Probability Model***

$\alpha = 0.792$

$\beta_1 = 0.038$

$\beta_2 = -0.079$

$\beta_3 = -0.148$

$\text{Pr(agreeing too little spent on education)} = 0.792 + 0.038x_1 - 0.079x_2 - 0.148x_1x_2$
